From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/equivalence.tex

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can shoe that these
syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
also locally-strict, or locally-posetal, because there is at most one 2-cell between compatible 1-cells.

@@

\begin{figure}
   \[
      \Tree [ [ {\tiny 1} {\tiny n} ] {\tiny m} ] ~\xrightarrow{}~
      \Tree [ {\tiny 1} [ {\tiny n} {\tiny m} ] ] ~\xrightarrow{}~
      \Tree [ [ {\tiny n} {\tiny m} ] {\tiny 1} ] ~\xrightarrow{}~
      \Tree [ [ {\tiny m} {\tiny n} ] {\tiny 1} ] ~\xrightarrow{}~
      \Tree [ {\tiny m} [ {\tiny n} {\tiny 1} ] ] ~\xrightarrow{}~
      \Tree [ {\tiny m} [ {\tiny 1} {\tiny n} ] ] ~
    \]
  \label{fig:plusplusswap}
  \caption{Braiding from transpositions, recursive case}
\end{figure}

@@

\begin{figure}
  \[
     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~\xrightarrow{\assoclp}~
     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow{\assocrp}~
     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~
   \]
 \label{fig:transpfrombraid}
 \caption{Transpositions from braiding}
\end{figure}

@@

\begin{definition}[${\gdenot{\blank}}$]
  \begin{gather*}
    \begin{aligned}
      \gdenot{\blank}_{0} & : U \to \UFin                                       \\
      \gdenot{\blank}_{0} & \defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \gdenot{\blank}_{0} & : X \iso Y \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
      \gdenot{\blank}_{1} & \defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

@@

Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal for both circuit is
computed to be, establishing their equivalence:

>>>>

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can show that these
syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
also locally-strict, or locally-posetal, because there is at most one 2-cell between compatible 1-cells.

@@

\begin{figure}
   \[
      \Tree [ [ {\tiny 1} {\tiny n} ] {\tiny m} ] ~\xrightarrow{}~
      \Tree [ {\tiny 1} [ {\tiny n} {\tiny m} ] ] ~\xrightarrow{}~
      \Tree [ [ {\tiny n} {\tiny m} ] {\tiny 1} ] ~\xrightarrow{}~
      \Tree [ [ {\tiny m} {\tiny n} ] {\tiny 1} ] ~\xrightarrow{}~
      \Tree [ {\tiny m} [ {\tiny n} {\tiny 1} ] ] ~\xrightarrow{}~
      \Tree [ {\tiny m} [ {\tiny 1} {\tiny n} ] ] ~
    \]
    \caption{Braiding from transpositions, recursive case}
    \label{fig:plusplusswap}
\end{figure}

@@

\begin{figure}
  \[
    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~\xrightarrow{\assoclp}~
    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow{\assocrp}~
    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~
  \]
  \caption{Transpositions from braiding}
  \label{fig:transpfrombraid}
\end{figure}

@@

\begin{definition}[${\gdenot{\blank}}$]
  \begin{gather*}
    \begin{aligned}
      \gdenot{\blank}_{0} & : U \to \UFin                                       \\
      \gdenot{\blank}_{0} & \defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
    \end{aligned}
    \qquad
    \begin{aligned}
      \gdenot{\blank}_{1} & : (c : X \iso Y) \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
      \gdenot{\blank}_{1} & \defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
    \end{aligned}
  \end{gather*}
\end{definition}

@@

Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both circuit is
computes to the following, establishing their equivalence:

### 2021-07-09 Pi+/paper/finite.tex

\begin{proposition}
  \label{prop:coxlongcox}
  The relations $\cox*$ and $\longcoxstar$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
  there is a $u$ such that $w \longcoxstar u$ and $v \longcoxstar u$.
\end{proposition}

>>>>

\begin{proposition}
  \label{prop:coxlongcox}
  The relations $\cox*$ and $\longcoxstar$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
  there is a $u$ such that $\coxcospan*{w}{v}{u}$.
\end{proposition}

### 2021-07-09 Pi+/paper/reversible1.tex
Fixes

The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the hardware
platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
permutations of finite sets provides a richer algebraic structure which we review in this section.

@@

These commutative rig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
equivalence to obtain a computational interpretation of the commutative rig
structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
groupoidification of a commutative rig, that is, a symmetric rig groupoid.

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,Carette2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

@@

\noindent The full set of level-2 combinators is large; the remaining combinators are listed in App.~\ref{app:leveltwo}.

\begin{toappendix}
  \label{app:leveltwo}

@@

>>>>

The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the
hardware platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an
``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical
model based on permutations of finite sets provides a richer algebraic structure, using which we describe a reversible
programming language in this section.

@@

These commutative rig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. To interpret natural number identities as reversible programs,
the logic needs to be equipped with values and types, and a notion of operational semantics and contextual equivalence,
giving a computational interpretation of the commutative rig structure~\cite{jamesInformationEffects2012}. On the
semantic side, the natural space to consider is the groupoidification of a commutative rig, that is, a symmetric rig
groupoid.

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

@@

\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo}.

\begin{toappendix}
  \subsection{2-combinators}

@@

  \label{app:leveltwo}

### 2021-07-09 Pi+/paper/ufin.tex

Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

@@

We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells
and work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
$\infty$-groupoids'' correspondence.

@@

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.

>>>>

In categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

@@

We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells.
We work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
$\infty$-groupoids'' correspondence.

@@

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.
